Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-10 2024pipit/section/01-Introduction.tex
tr

\begin{itemize}
  \item we describe a key issue with prior program logics for Lustre-style languages (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified rely-guarantee contracts (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove preservation of the dynamic (bigstep) semantics to an abstract transition system suitable for verifying properties (\autoref{s:transition:proof});
  \item similarly, we identify the invariants and lemmas required to prove that the executable transition system preserves the original semantics (\autoref{s:extraction:proof});
  \item we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).
\end{itemize}

% \begin{itemize}
%   \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
%   \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
%   \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
%   % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
%   \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
% \end{itemize}

%CUT
% \begin{itemize}
%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
%     % TODO: CUT WAFFLE
%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
% \end{itemize}

% \begin{itemize}
%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
% \end{itemize}

>>>>

\begin{itemize}
  \item we motivate our work, which allows us to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified rely-guarantee contracts (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove preservation of the dynamic semantics to an abstract transition system suitable for verifying properties (\autoref{s:transition:proof});
  \item similarly, we identify the invariants and lemmas required to prove that the executable transition system preserves the original semantics (\autoref{s:extraction:proof});
  \item finally, we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).
\end{itemize}

### 2024-01-10 2024pipit/section/30-Core.tex
tr

The causality restriction gives us some important properties about the metatheory.
The most important property is that the dynamic semantics form a total function: given a streaming history and a causal expression, we can evaluate the expression to a value.

@@

The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
An alternative semantics would be to have the environment outside the semantics invent a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
The above substitution theorem can be used to show that these two semantics are equivalent.
Thanks to causality, we can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
This fact can be used to ``seed'' evaluation by starting with an arbitrary value:
\begin{theorem}[bigstep-rec-causal]
  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x := v']}{e}{v})$.
\end{theorem}

>>>>

The causality restriction gives us some important properties about the metatheory.
The most important property is that the dynamic semantics form a total function: given a streaming history and a causal expression, we can evaluate the expression to a value.
These properties are mechanised in \fstar{}.

@@

The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
An alternative semantics would be to have the environment outside the semantics invent a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
The above substitution theorem can be used to show that these two semantics are equivalent.
Thanks to causality, we can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
This fact can be used to ``seed'' evaluation by starting with an arbitrary value:
\begin{theorem}[bigstep-rec-causal]
  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x \mapsto v']}{e}{v})$.
\end{theorem}

### 2024-01-10 2024pipit/section/35-Primitive-Tables.tex
ok
tr

\subsection{Primitive tables}
\label{s:core:primitive-tables}

The core language presented in \autoref{f:core-grammar} has a restricted set of primitives.
Our actual implementation is parameterised by the set of primitive types and operations.

In addition to a standard deeply-embedded primitive table which supports integers, tuples, booleans and reals, we also provide a shallowly-embedded primitive table, where operations can be arbitrary pure \fstar{} functions.
This shallow embedding also allows straightforward integration of \fstar{} types and functions, as well as proofs inside streaming functions.

This parameterisation serves mainly as a convenience.
For the formalisation presented here, we assume a simple static set of primitives.

In our current implementation, we cannot encode primitives that have preconditions, as the preconditions are checked eagerly by \fstar{}, before anything is known about the surrounding streaming values.

To work around this limitation, we encode certain primitive preconditions as implications on the postcondition.
For example, to encode the array indexing operation in \autoref{s:motivation}, we use an unsafe variant of the indexing function:
\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @val@ Array.index_unsafe ($\textit{array}: \mbox{const_array}~\alpha$) ($\textit{index}$: $\NN$): \\
  \> $v: \alpha \{ \textit{index} < \mbox{Array.length}~\textit{array} \implies v = \mbox{Array.index}~\textit{array}~\textit{index} \}$
\end{tabbing}

By wrapping this unsafe index function inside a contract that requires the index to be valid, we can prove the postcondition on the streaming level.

In the future, we plan to investigate allowing preconditions on non-streaming functions, so that safe variants can be used directly.

>>>>

% KILL I don't think this is really useful

% \subsection{Primitive tables}
% \label{s:core:primitive-tables}

% The core language presented in \autoref{f:core-grammar} has a restricted set of primitives.
% Our actual implementation is parameterised by the set of primitive types and operations.

% In addition to a standard deeply-embedded primitive table which supports integers, tuples, booleans and reals, we also provide a shallowly-embedded primitive table, where operations can be arbitrary pure \fstar{} functions.
% This shallow embedding also allows straightforward integration of \fstar{} types and functions, as well as proofs inside streaming functions.

% This parameterisation serves mainly as a convenience.
% For the formalisation presented here, we assume a simple static set of primitives.

### 2024-01-10 2024pipit/section/40-Transition.tex
ok
tr

\section{Transition systems}
\label{s:transition}
\input{figures/system-types.tex}

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input environment and the result type.
It also contains two internal environments: firstly, the state environment describes the private state required to execute the machine; secondly, the free environment contains any extra input values that the transition system would like to quantify over.
The free environment is used to allow the system to concoct arbitrary values, when it would not otherwise be able to return a concrete value.
For contracts which abstract over the implementation, the natural definition of the step function would involve an existential quantifier: there exists some value that satisfies the specification.
Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation with existential quantifiers complicates the resulting transition, as the quantifiers get between the value and its definition.
Instead, the free environment provides the step function with a fresh unconstrained value of the desired type, and the step function itself can constrain it however it wishes.

The free environment is used to generate similar systems to the transition systems generated by existing systems such as Kind2 \cite{champion2016kind2}, which can generate fresh variables at will.
The main distinction is that embedding such a translation inside a theorem prover and proving it correct requires some ingenuity.

@@

@@

\subsection{Translation correctness proofs}

>>>>

\section{Abstract transition systems}
\label{s:transition}
\input{figures/system-types.tex}

To prove properties about Pipit programs, we translate to an \emph{abstract} transition system, so-called because it abstracts away the implementation details of contract instantiations.
For extraction we also translate to \emph{executable} transition systems, which we discuss in \autoref{s:extraction}.

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input context and the result type.
It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to quantify over.
The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.

For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: there exists some value that satisfies the specification.
Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block normalisation and result in a more complex transition system.
Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.

As usual, the step-result contains the updated state for the transition system, as well as the result value.
The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
For the transition system corresponding to an expression $e$, these propositions are analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.

% There is a slight difference with how followed-by delays are handled --- the checked semantics checks the delayed expression in the delayed context, while the transition system checks one step ahead ---

% The free context is used to generate similar systems to the transition systems generated by existing systems such as Kind2 \cite{champion2016kind2}, which can generate fresh variables at will.
% The main distinction is that embedding such a translation inside a theorem prover and proving it correct requires some ingenuity.

@@

\autoref{f:system-translation-contexts} defines the internal state and free contexts required for an expression.
For most expression forms, the state and free contexts are defined by taking the union of the contexts of sub-expressions.
Followed-by delays introduce a local state variable $x$, in which to store the most recent stream value.
We generate a fresh variable here, though the implementation uses de Bruijn indices.
Recursive streams and contracts both introduce new bindings into the free context, assuming that their binders $x$ are fresh.

@@

\subsection{Translation correctness proofs}
\label{s:transition:proof}

### 2024-01-10 2024pipit/section/60-Evaluation.tex
ok
tr

@@

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\error{\textit{error}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}
\pagebreak

Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
As with the previous examples, the two properties can be automatically verified.
% by \fstar{}'s proof automation.
Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.

>>>>

\TODO{EVALUATION NEEDS DOING:}
\begin{itemize}
  \item describe time-triggered CAN implementation
  \item whole implementation includes the extended version of trigger logic
  \item verified machine arithmetic
  \item for which we can generate real-time C code
  \item describe coffee machine controller
  \item simple, but fully verified according to an abstract spec
  \item and for which we generate real-time C code
\end{itemize}

@@

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\error{\textit{error}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}

Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
As with the previous examples, the two properties can be automatically verified.
% by \fstar{}'s proof automation.
Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \COFFEEURL}.
